---
title: "Portland Air Quality Data Mining Project"
author: "Jennifer Scott, Thammik Leungpathomaram"
date: "2026-02-13"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# 1. Introduction

This project uses daily air quality and weather observations for Portland (2019–2021) to:

1. Predict daily ozone concentration (regression)

2. Classify unhealthy air quality days using AQI categories (classification)

3. Evaluate models using a time-based split (train: 2019–2020, test: 2021)

4. Explore seasonality using autocorrelation and Fourier terms

5. Apply PCA to address multicollinearity among weather variables (dimension reduction)

Our primary goal is interpretability and sound evaluation rather than maximizing the number of algorithms implemented.


# 2.Library Block

```{r library block}
library(utils)
library(ggplot2)
library(dplyr)
library(caret)
library(corrplot)
library(tidyr)
library(leaps)
library(forecast)
library(zoo)
```

# 3. Data Loading

```{r data loading}
source("config.R")
aqi <- read.csv(csv_path)
```


```{r}
#str(aqi)
#summary(aqi)
```

# 4. Missing Data Review and Cleaning Strategy

## 4.1 Missingness summary

We begin by summarizing missing values for each field. This helps us decide which variables require imputation and whether any fields should be excluded due to excessive missingness.

```{r missing data summary}
missing_summary <- aqi %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  mutate(pct_missing = n_missing / nrow(aqi) * 100) %>%
  arrange(desc(n_missing))

missing_summary

aqi$Date<-as.Date(aqi$Date)

# impute<-FALSE # I think we should remove this
```

## 4.2 Handling missing target values (Ozone_ppm)
We do not impute the target variable (ozone). Rows missing the target are removed from modeling datasets.

```{r missing target values}

# Drop records that are missing the target variable, as imputation would be unreliable and could bias the model.
aqi_clean <- aqi %>%
  filter(!is.na(Ozone_ppm))
```


## 4.3 Impute remaining numeric predictors

For predictor variables, we impute missing numeric values using a simple strategy.

```{r impute remaining predictors}

impute_method <- "median" # mean can be used as well

impute_fun <- if (impute_method == "median") median else mean

aqi_clean <- aqi_clean %>%
  mutate(across(where(is.numeric),
                ~ifelse(is.na(.), impute_fun(., na.rm = TRUE), .)))

# Quick check of the missing days. 29 days, consisting of stretches between 3-12 days.
#missing_days <-aqi[is.na(aqi$Ozone_ppm), ]$Date

```

# 5. Exploratory Data Analysis

## 5.1 Ozone distribution

```{r}
ggplot(aqi_clean, aes(x = Ozone_ppm)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  labs(title = "Distribution of Ozone (ppm)", x = "Ozone (ppm)", y = "Count") +
  theme_minimal()

summary(aqi_clean$Ozone_ppm)
```

```{r}
ggplot(aqi_clean, aes(y = Ozone_ppm)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(title = "Ozone Distribution - Outlier Check", y = "Ozone (ppm)") +
  theme_minimal()
```

## 5.2 Ozone over time

This plot helps identify long-term trend and year-to-year differences.

```{r}

# dot plot
ggplot(aqi_clean %>% filter(!is.na(Ozone_ppm)), 
       aes(x = Date, y = Ozone_ppm, color = as.factor(Year))) +
  geom_point(alpha = 0.6) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  labs(title = "Ozone Levels Over Time (2019-2021)",
       x = "Date", y = "Ozone (ppm)", color = "Ozone Level") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# line chart

aqi_clean <- aqi_clean %>%
  mutate(Year = lubridate::year(Date))

ggplot(aqi_clean, aes(x = Date, y = Ozone_ppm, color = as.factor(Year))) +
  geom_line(alpha = 0.6) +
  geom_smooth(aes(group = 1), method = "loess", se = TRUE, alpha = 0.2) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  labs(title = "Ozone Levels Over Time (2019–2021)",
       x = "Date", y = "Ozone (ppm)", color = "Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## 5.3 Seasonality (Day of Year)

```{r}


ggplot(aqi_clean, aes(x = DayOfYear, y = Ozone_ppm)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(formula = y~x, method = "loess", color = "red", se = TRUE) +
  scale_x_continuous(breaks = c(1, 91, 182, 274, 365),
                     labels = c("Jan 1", "Apr 1", "Jul 1", "Oct 1", "Dec 31")) +
  labs(title = "Seasonal Pattern in Ozone Levels",
       x = "Day of Year", y = "Ozone (ppm)") +
  theme_minimal()

```

```{r}
ggplot(aqi_clean, aes(x = as.factor(Month), y = Ozone_ppm)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(title = "Ozone Distribution by Month",
       x = "Month", y = "Ozone (ppm)") +
  theme_minimal()
```

# 6. Train/Test Split (Time Based)

To prevent “peeking into the future,” we split chronologically:

- Train: 2019–2020

- Test: 2021

```{r split data}

train_data <- aqi_clean %>% filter(Year %in% c(2019, 2020))
test_data  <- aqi_clean %>% filter(Year == 2021)

dim(train_data)
dim(test_data)

```

# 7. Regression Modeling (Predict Ozone)

## 7.1 Define predictors (avoid leakage)

We avoid variables that are direct transformations of the target or derived from it.

- Ozone_ppb is a deterministic transform of Ozone_ppm

- AQI and AQI_Ozone can include ozone information and cause leakage

```{r define predictors}
all_numeric <- train_data %>% select(where(is.numeric)) %>% names()

leakage_vars <- c("Ozone_ppm", "Ozone_ppb", "AQI", "AQI_Ozone")
predictors <- setdiff(all_numeric, leakage_vars)

# Optional: if wee need to remove additional fields
# predictors <- setdiff(predictors, c("AQI_PM25"))

```

## 7.2 Correlation review (drop zero-variance columns first)

```{r correlation review}

lr_data <- train_data %>%
  select(Ozone_ppm, all_of(predictors))

nzv <- caret::nearZeroVar(lr_data)
lr_data2 <- lr_data[, -nzv, drop = FALSE]

cor_matrix <- cor(lr_data2, use = "pairwise.complete.obs")
corrplot(cor_matrix)

# high_cor <- which(abs(cor_matrix) > 0.7, arr.ind = TRUE)
# high_cor_pairs <- data.frame(
#   var1 = rownames(cor_matrix)[high_cor[,1]],
#   var2 = colnames(cor_matrix)[high_cor[,2]],
#   correlation = cor_matrix[high_cor]
# ) %>%
#   distinct() %>%
#   filter(var1!=var2) %>%
#   # Removes 'duplicate' pairs
#   filter(!duplicated(paste(pmin(var1, var2), pmax(var1, var2)))) %>%
#   arrange(desc(abs(correlation)))
# high_cor_pairs
# 
# corrplot(cor_matrix)

```

## 7.3 Baseline linear regression

```{r linear regression}

formula_full <- as.formula(paste("Ozone_ppm ~", paste(predictors, collapse = " + ")))
lm_full <- lm(formula_full, data = train_data)
summary(lm_full)

cat("Train Adj R^2:", summary(lm_full)$adj.r.squared, "\n")

```

Evaluate on 2021 data:

```{r lm on 2021}

pred_lm <- predict(lm_full, test_data)
obs_lm  <- test_data$Ozone_ppm

rmse_lm <- RMSE(pred_lm, obs_lm)
mae_lm  <- MAE(pred_lm, obs_lm)

ss_res <- sum((obs_lm - pred_lm)^2)
ss_tot <- sum((obs_lm - mean(obs_lm))^2)
r2_lm <- 1 - ss_res / ss_tot

cat("Linear Regression (Test 2021): RMSE =", rmse_lm,
    " MAE =", mae_lm,
    " R^2 =", r2_lm, "\n")

```

## 7.4 Reduced linear regression (parsimony)

A reduced model improves interpretability and may generalize better.

```{r parsimony lm}

significant_vars <- c("NO2_ppb_or_ppm", "SO2_ppb_or_ppm", 
                      "Wind2m_ms", "DewPoint_C", "Precip_mm", "Solar_Wm2", "DayOfYear")

formula_reduced <- as.formula(paste("Ozone_ppm ~", paste(significant_vars, collapse = " + ")))
lm_reduced <- lm(formula_reduced, data = train_data)
summary(lm_reduced)

pred_red <- predict(lm_reduced, test_data)
obs_red  <- test_data$Ozone_ppm

cat("Reduced LM (Test 2021): RMSE =", RMSE(pred_red, obs_red),
    " MAE =", MAE(pred_red, obs_red), "\n")

```

# Stepwise 

```{r}
trControl <- trainControl(method = "none")

# Stepwise selection 
model_stepwise <- train(
  formula_full,
  data = train_data,
  trControl = trControl,
  method = "glmStepAIC",
  direction = "both",
  trace = FALSE
)
```

```{r}
lm_stepwise <- model_stepwise$finalModel
```


```{r}

lm_stepwise_r2 <- 1 - (lm_stepwise$deviance / lm_stepwise$null.deviance)
n <- nrow(train_data)
p <- length(coef(lm_stepwise)) - 1
lm_stepwise_adj_r2 <- 1 - (1 - lm_stepwise_r2) * (n - 1) / (n - p - 1)
summary(lm_stepwise)
cat("Train Adj R^2:", lm_stepwise_adj_r2, "\n")

pred <- predict(lm_stepwise, test_data)
actual <- test_data$Ozone_ppm
ss_res <- sum((actual - pred)^2)
ss_tot <- sum((actual - mean(actual))^2)
test_r2 <- 1 - ss_res / ss_tot
n_test <- nrow(test_data)
test_adj_r2 <- 1 - (1 - test_r2) * (n_test - 1) / (n_test - p - 1)
cat("Test Adj R^2: ", test_adj_r2, "\n")


```
With Imputation:
Train Adj R^2: 0.5937668 
Test Adj R^2:  0.2704882 


Without Imputation:
Train Adj R^2: 0.6280177 
Test Adj R^2:  0.281659 
Only removes 1 predictor... Forward and backward have similar performance. Tried using best subset selection instead, but there's way too many predictors for that.

62% is decent, but 11 predictors is a lot.


Try removing predictors where p >0.05. (CO_ppm and Pressure_kPa)
```{r}
# removed these two for now "PM25_ugm3", "PM10_ugm3", but may need to add these back in. R2 is much worse without them 

significant_vars <- c("NO2_ppb_or_ppm", "SO2_ppb_or_ppm", 
                      "Wind2m_ms", "DewPoint_C", "Precip_mm", "Solar_Wm2", "DayOfYear")

formula_reduced <- as.formula(paste("Ozone_ppm ~", paste(significant_vars, collapse = " + ")))
lm_reduced <- lm(formula_reduced, data = train_data)
summary(lm_reduced)

cat("Train Adj R^2:", summary(lm_reduced)$adj.r.squared, "\n")

pred <- predict(lm_reduced, test_data)
actual <- test_data$Ozone_ppm
ss_res <- sum((actual - pred)^2)
ss_tot <- sum((actual - mean(actual))^2)
test_r2 <- 1 - ss_res / ss_tot
n_test <- nrow(test_data)
p <- length(coef(lm_reduced)) - 1
test_adj_r2 <- 1 - (1 - test_r2) * (n_test - 1) / (n_test - p - 1)
cat("Test Adj R^2: ", test_adj_r2, "\n")
```

With Imputation:
Train Adj R^2: 0.5363423 
Test Adj R^2:  0.3739211 

Without:

Train Adj R^2: 0.5819515 
Test Adj R^2:  0.2246784 


# 8. Seasonality and Fourier Time_Series Regression

## 8.1 Autocorrelation (ACF)

```{r autocorrelation}

ozone_train_ts <- ts(train_data$Ozone_ppm, frequency = 365)
Acf(ozone_train_ts, main = "ACF of Daily Ozone (Train: 2019–2020)")

```

A slowly decaying ACF and repeating structure suggests strong seasonality. For ozone, this is expected due to sunlight-driven chemistry and summer weather patterns.


## 8.2 Select Fourier comlexity (K)

We use AIC to select the number of Fourier harmonics (K). Larger K captures more complex seasonal shape but risks overfitting.

```{r fourier}

 k_values <- 1:10
 aic_values <- sapply(k_values, function(k) {
   formula <- as.formula(paste("ozone_train_ts ~ trend + fourier(ozone_train_ts, K=", k, ") +",
                                paste(significant_vars, collapse = " + ")))
   fit <- tslm(formula, data = train_data)
   return(AIC(fit))
 })

  ggplot(data.frame(K = k_values, AIC = aic_values), aes(x = K, y = AIC)) +
   geom_line() +
   geom_point() +
   scale_x_continuous(breaks = k_values) +
   labs(title = "AIC vs Number of Fourier Terms (K)", x = "K", y = "AIC") +
   theme_minimal()
```
Based on the plot K=3 or 5 is ideal.

```{r}
# tslm without predictors
# very weak results (adjr^2=0.286)

#ozone.tslm <- tslm(ozone.ts ~ trend + fourier(ozone.ts, K=3))
#summary(ozone.tslm)
```

```{r}

# fourier tslm with predictors.
formula1 <- as.formula(paste("ozone_train_ts ~ trend + fourier(ozone_train_ts, K=3) +",
                             paste(predictors, collapse = " + ")))

formula2 <- as.formula(paste("ozone_train_ts ~ trend + fourier(ozone_train_ts, K=3) +",
                             paste(significant_vars, collapse = " + ")))

ozone.tslm.fourier1 <- tslm(formula1, data = train_data)
ozone.tslm.fourier2 <- tslm(formula2, data = train_data)

#summary(ozone.tslm.fourier1)
summary(ozone.tslm.fourier1)$adj.r.squared
summary(ozone.tslm.fourier2)$adj.r.squared

```
With imputation: Adj R^2 of 66.6

Without imputation: Adjusted R^2 of 70.45% with K=5. Strong results, although this model has 22 predictors.

## 8.3 Fit Fourier + predictors and forecast 2021 properly

This is the corrected out-of-sample evaluation. We:

- compute Fourier terms on the training series

- project Fourier terms forward into the test horizon using fourierf()

- forecast with forecast(..., xreg=...)

```{r}
# --- Forecast out-of-sample properly (Fourier + xreg)
K <- 3
h <- nrow(test_data)

# time series used to fit
ozone_train_ts <- ts(train_data$Ozone_ppm, frequency = 365)

# Build xreg matrices from predictors (must match train/test row counts)
xreg_train <- train_data %>%
  select(all_of(predictors)) %>%
  mutate(across(everything(), as.numeric)) %>%
  as.matrix()

xreg_test <- test_data %>%
  select(all_of(predictors)) %>%
  mutate(across(everything(), as.numeric)) %>%
  as.matrix()

# Fourier terms
four_train <- fourier(ozone_train_ts, K = K)
four_test  <- fourierf(ozone_train_ts, K = K, h = h)

xreg_train_all <- cbind(four_train, xreg_train)
xreg_test_all  <- cbind(four_test,  xreg_test)

# Fit + forecast
fit_fourier <- tslm(ozone_train_ts ~ trend + season, xreg = xreg_train_all)
fc <- forecast(fit_fourier, xreg = xreg_test_all, h = h)

# Accuracy vs actual
accuracy(fc$mean, test_data$Ozone_ppm)

```

# 9. classification Modeling (Logistic Regression)

## 9.1 Define classification target

We classify days as “Orange or worse” using the AQI indicator already in the dataset:

- AQI_AtLeast_Orange = 1 if AQI ≥ 101

```{r logit}
train_cls <- train_data %>%
  mutate(AQI_AtLeast_Orange = factor(AQI_AtLeast_Orange, levels = c(0, 1)))

test_cls <- test_data %>%
  mutate(AQI_AtLeast_Orange = factor(AQI_AtLeast_Orange, levels = c(0, 1)))

```

## 9.2 Logistic regression (interpretable classifier)
We use a small, interpretable set of predictors (similar to regression drivers).

```{r}

cls_vars <- c("NO2_ppb_or_ppm", "SO2_ppb_or_ppm", "Wind2m_ms",
              "DewPoint_C", "Precip_mm", "Solar_Wm2", "DayOfYear")

cls_formula <- as.formula(paste("AQI_AtLeast_Orange ~", paste(cls_vars, collapse = " + ")))
logit_model <- glm(cls_formula, data = train_cls, family = binomial())
summary(logit_model)

```

# 10. PCA (Dimension Reduction)

PCA is used here to reduce redundancy among correlated weather variables. We:

- standardize predictors

- compute PCs

- examine variance explained and loadings

- optionally refit regression using PCs

## 10.1 Prepare PCA predictors

```{r}

pca_vars <- c("Temp_C","TempMax_C","TempMin_C","Wind2m_ms","Wind10m_ms",
              "RH_percent","DewPoint_C","Precip_mm","Pressure_kPa","Solar_Wm2")

pca_train <- train_data %>%
  select(all_of(pca_vars)) %>%
  mutate(across(everything(), as.numeric))

pca_fit <- prcomp(pca_train, center = TRUE, scale. = TRUE)

# Scree plot
plot(pca_fit, type = "l", main = "Scree Plot (PCA)")
summary(pca_fit)

```

## 10.2 Interpret component loadings

```{r}

loadings <- pca_fit$rotation
head(loadings)

```

## 10.3 Regression using PCs
We use enough PCs to explain ~80-90% variance.

```{r}

var_explained <- cumsum(pca_fit$sdev^2) / sum(pca_fit$sdev^2)
n_pc <- which(var_explained >= 0.85)[1]
n_pc

```

```{r}

train_pcs <- as.data.frame(pca_fit$x[, 1:n_pc, drop = FALSE])
train_pcs$Ozone_ppm <- train_data$Ozone_ppm

pca_model <- lm(Ozone_ppm ~ ., data = train_pcs)
summary(pca_model)

```

Evaluate on 2021 data:

```{r}

pca_test <- test_data %>%
  select(all_of(pca_vars)) %>%
  mutate(across(everything(), as.numeric))

test_scores <- predict(pca_fit, newdata = pca_test)[, 1:n_pc, drop = FALSE]
pred_pca <- predict(pca_model, newdata = as.data.frame(test_scores))

cat("PCA Regression (Test 2021): RMSE =", RMSE(pred_pca, test_data$Ozone_ppm),
    " MAE =", MAE(pred_pca, test_data$Ozone_ppm), "\n")

```


# Don't what to do with these sections below just yet
Normal tlsm
```{r}
train_monthly <- train_data %>%
  group_by(Year, Month) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE))

test_monthly <- test_data %>%
  group_by(Year, Month) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE))

ozone.ts.monthly <- ts(train_monthly$Ozone_ppm, frequency = 12, start = c(2019, 1))

#formula1 <- as.formula(paste("ozone.ts.monthly ~ trend + season +", paste(predictors, collapse = " + ")))
ts_vars <- c("NO2_ppb_or_ppm", "SO2_ppb_or_ppm", "Wind2m_ms", 
                      "DewPoint_C", "Precip_mm", "Solar_Wm2")

formula2 <- as.formula(paste("ozone.ts.monthly ~ trend + season +", paste(ts_vars, collapse = " + ")))

#ozone.tslm.monthly1 <- tslm(formula1, data = train_monthly)
ozone.tslm.monthly2 <- tslm(formula2, data = train_monthly)

#cat("Train Adj R^2 (formula1):", summary(ozone.tslm.monthly1)$adj.r.squared, "\n")
cat("Train Adj R^2 (formula2):", summary(ozone.tslm.monthly2)$adj.r.squared, "\n")

```


```{r}
test.ts.monthly <- ts(test_monthly$Ozone_ppm, frequency = 12, start = c(2021, 1))

#forecast1 <- forecast(ozone.tslm.monthly1, newdata = test_monthly)
forecast2 <- forecast(ozone.tslm.monthly2, newdata = test_monthly[, significant_vars])

#cat("Train Adj R^2 (formula1):", summary(ozone.tslm.monthly1)$adj.r.squared, "\n")
cat("Train Adj R^2 (formula2):", summary(ozone.tslm.monthly2)$adj.r.squared, "\n")

#accuracy(forecast1, test.ts.monthly)
accuracy(forecast2, test.ts.monthly)
```

Predictor-less ARIMA
```{r}

# ozone.arima <- auto.arima(ozone_train_ts)
# summary(ozone.arima)

```

ARIMA with Predictors
```{r}

xreg_matrix <- as.matrix(train_data[, c("NO2_ppb_or_ppm", "CO_ppm", "SO2_ppb_or_ppm", 
                                         "Temp_C", "Wind2m_ms", "RH_percent", 
                                         "DewPoint_C", "Precip_mm", "Pressure_kPa", 
                                         "Solar_Wm2", "DayOfYear")])

ozone.arima.xreg <- auto.arima(ozone_train_ts, xreg = xreg_matrix)
summary(ozone.arima.xreg)
```
using predictors somehow makes ARIMA worse. 
