---
title: "Portland Air Quality Data Mining Project"
author: "Jennifer Scott, Thammik Leungpathomaram"
date: "2026-02-13"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

```

# 1. Introduction

This project uses daily air quality and weather observations for Portland (2019–2021) to:

1. Predict daily ozone concentration (regression)

2. Classify unhealthy air quality days using AQI categories (classification)

3. Evaluate models using a time-based split (train: 2019–2020, test: 2021)

4. Explore seasonality using autocorrelation and Fourier terms

5. Apply PCA to address multicollinearity among weather variables (dimension reduction)

Our primary goal is interpretability and sound evaluation rather than maximizing the number of algorithms implemented. Understanding the relationships between meteorological conditions and ozone levels is key for public health insights and potential forecasting applications.conditions and ozone variability will help inform forecasting and public health alerts. 


# 2.Library Block

```{r message=FALSE, warning=FALSE}
library(utils)
library(ggplot2)
library(dplyr)
library(caret)
library(corrplot)
library(tidyr)
library(leaps)
library(forecast)
library(zoo)
```

# 3. Data Loading

```{r data loading}
source("config.R")
aqi <- read.csv(csv_path)
```


```{r}
#str(aqi)
#summary(aqi)
```

# 4. Missing Data Review and Cleaning Strategy

## 4.1 Missingness summary

We begin by summarizing missing values for each field. This helps us decide which variables require imputation and whether any fields should be excluded due to excessive missingness. Because the modeling goal is prediction rather than caussal inference, simple imputation strategies like median are sufficient and prevent the loss of potentially useful observations. We do not impute the target variable (ozone) because this could bias the model; instead, we remove rows missing the target from the modeling datasets.

```{r missing data summary}
missing_summary <- aqi %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  mutate(pct_missing = n_missing / nrow(aqi) * 100) %>%
  arrange(desc(n_missing))

missing_summary

aqi$Date<-as.Date(aqi$Date)

```

## 4.2 Handling missing target values (Ozone_ppm)
We do not impute the target variable (ozone). Rows missing the target are removed from modeling datasets.

```{r missing target values}

# Drop records that are missing the target variable, as imputation would be unreliable and could bias the model.
aqi_clean <- aqi %>%
  filter(!is.na(Ozone_ppm))
```


## 4.3 Impute remaining numeric predictors

For predictor variables, we impute missing numeric values using a simple strategy.

```{r impute remaining predictors}

impute_method <- "median" 

impute_fun <- if (impute_method == "median") median else mean

aqi_clean <- aqi_clean %>%
  mutate(across(where(is.numeric),
                ~ifelse(is.na(.), impute_fun(., na.rm = TRUE), .)))

# Quick check of the missing days. 29 days, consisting of stretches between 3-12 days.
#missing_days <-aqi[is.na(aqi$Ozone_ppm), ]$Date

```

# 5. Exploratory Data Analysis

## 5.1 Ozone distribution

```{r}
ggplot(aqi_clean, aes(x = Ozone_ppm)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  labs(title = "Distribution of Ozone (ppm)", x = "Ozone (ppm)", y = "Count") +
  theme_minimal()

summary(aqi_clean$Ozone_ppm)
```

```{r}
ggplot(aqi_clean, aes(y = Ozone_ppm)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(title = "Ozone Distribution - Outlier Check", y = "Ozone (ppm)") +
  theme_minimal()
```



## 5.2 Ozone over time

This plot helps identify long-term trend and year-to-year differences.

```{r message=FALSE}


# line chart

aqi_clean <- aqi_clean %>%
  mutate(Year = lubridate::year(Date))

ggplot(aqi_clean, aes(x = Date, y = Ozone_ppm, color = as.factor(Year))) +
  geom_line(alpha = 0.6) +
  geom_smooth(aes(group = 1), method = "loess", formula = y ~ x, se = TRUE, alpha = 0.2) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  labs(title = "Ozone Levels Over Time (2019–2021)",
       x = "Date", y = "Ozone (ppm)", color = "Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# dot plot
ggplot(aqi_clean %>% filter(!is.na(Ozone_ppm)), 
       aes(x = Date, y = Ozone_ppm, color = as.factor(Year))) +
  geom_point(alpha = 0.6) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  labs(title = "Ozone Levels Over Time (2019-2021)",
       x = "Date", y = "Ozone (ppm)", color = "Ozone Level") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```





## 5.3 Seasonality (Day of Year)

```{r}


ggplot(aqi_clean, aes(x = DayOfYear, y = Ozone_ppm)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(formula = y~x, method = "loess", color = "red", se = TRUE) +
  scale_x_continuous(breaks = c(1, 91, 182, 274, 365),
                     labels = c("Jan 1", "Apr 1", "Jul 1", "Oct 1", "Dec 31")) +
  labs(title = "Seasonal Pattern in Ozone Levels",
       x = "Day of Year", y = "Ozone (ppm)") +
  theme_minimal()

```

```{r}
ggplot(aqi_clean, aes(x = as.factor(Month), y = Ozone_ppm)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(title = "Ozone Distribution by Month",
       x = "Month", y = "Ozone (ppm)") +
  theme_minimal()
```

The seasonal pattern is clear: ozone concentrations increase during late spring and summer months and decline during winter. This is consistent with atmospheric chemistry, where ozone formation is driven by sunlight and temperature-dependent reactions involving nitrogen oxides and volatile organic compounds. The boxplot by month also confirms higher median ozone levels in late spring and summermonths (April–August) compared to Fall / winter months (October–January). It is also interesting to note the the inner quartile is much wider during the fall / winter months than the higher median months of spring / summer.

# 6. Train/Test Split (Time Based)

To prevent “peeking into the future,” we split chronologically:

- Train: 2019–2020

- Test: 2021

According to "Machine Learning for Business," models are best trained on earlier data and evaluated on later data to simulate real-world forecasting. This approach avoids data leakage and provides a more realistic assessment of model performance on unseen future data. 

```{r split data}

train_data <- aqi_clean %>% filter(Year %in% c(2019, 2020))
test_data  <- aqi_clean %>% filter(Year == 2021)

dim(train_data)
dim(test_data)

```

# 7. Regression Modeling (Predict Ozone)

## 7.1 Define predictors (avoid leakage)

We avoid variables that are direct transformations of the target or derived from it.

- Ozone_ppb is a deterministic transform of Ozone_ppm

- AQI and AQI_Ozone can include ozone information and cause leakage

```{r define predictors}
all_numeric <- train_data %>% select(where(is.numeric)) %>% names()

leakage_vars <- c("Ozone_ppm", "Ozone_ppb", "AQI", "AQI_Ozone")
predictors <- setdiff(all_numeric, leakage_vars)

```

## 7.2 Correlation review 

```{r correlation review}

lr_data <- train_data %>%
  select(Ozone_ppm, all_of(predictors))

nzv <- caret::nearZeroVar(lr_data)
lr_data2 <- lr_data[, -nzv, drop = FALSE]

cor_matrix <- cor(lr_data2, use = "pairwise.complete.obs")
corrplot(cor_matrix)

# high_cor <- which(abs(cor_matrix) > 0.7, arr.ind = TRUE)
# high_cor_pairs <- data.frame(
#   var1 = rownames(cor_matrix)[high_cor[,1]],
#   var2 = colnames(cor_matrix)[high_cor[,2]],
#   correlation = cor_matrix[high_cor]
# ) %>%
#   distinct() %>%
#   filter(var1!=var2) %>%
#   # Removes 'duplicate' pairs
#   filter(!duplicated(paste(pmin(var1, var2), pmax(var1, var2)))) %>%
#   arrange(desc(abs(correlation)))
# high_cor_pairs
# 
# corrplot(cor_matrix)

```
Outside of expect collinear groups(i.e. temperature or time related predictors), there is moderate correlation between sunlight, relative humidity, and temperature. There is also weak correlation between NO2, SO2, and CO2.
## 7.3 Baseline linear regression

A baseline model using all possible predictors.

```{r linear regression}

formula_full <- as.formula(paste("Ozone_ppm ~", paste(predictors, collapse = " + ")))
lm_full <- lm(formula_full, data = train_data)
summary(lm_full)

lm_full_test_pred <- predict(lm_full, newdata = test_data)

print("Training:")
cat("R^2:", summary(lm_full)$adj.r.squared, "\n")
print(accuracy(lm_full$fitted.values, train_data$Ozone_ppm))

```

Evaluate on 2021 data:

```{r lm on 2021}

pred_lm <- predict(lm_full, test_data)
obs_lm  <- test_data$Ozone_ppm

rmse_lm <- RMSE(pred_lm, obs_lm)
mae_lm  <- MAE(pred_lm, obs_lm)

ss_res <- sum((obs_lm - pred_lm)^2)
ss_tot <- sum((obs_lm - mean(obs_lm))^2)
r2_lm <- 1 - ss_res / ss_tot

cat("Linear Regression (Test 2021): RMSE =", rmse_lm,
    " MAE =", mae_lm,
    " R^2 =", r2_lm, "\n")

```

## 7.4 Reduced linear regression (parsimony)

A reduced model improves interpretability and may generalize better.

```{r parsimony lm}

significant_vars <- c("NO2_ppb_or_ppm", "SO2_ppb_or_ppm", 
                      "Wind2m_ms", "DewPoint_C", "Precip_mm", "Solar_Wm2", "DayOfYear")

formula_reduced <- as.formula(paste("Ozone_ppm ~", paste(significant_vars, collapse = " + ")))
lm_reduced <- lm(formula_reduced, data = train_data)
summary(lm_reduced)

pred_red <- predict(lm_reduced, test_data)
obs_red  <- test_data$Ozone_ppm

cat("Reduced LM (Test 2021): RMSE =", RMSE(pred_red, obs_red),
    " MAE =", MAE(pred_red, obs_red), "\n")

cat("Train RMSE:", accuracy(lm_reduced$fitted.values, train_data$Ozone_ppm)[, "RMSE"], "\n")
cat("Test RMSE: ", accuracy(pred_red, obs_red)[, "RMSE"], "\n")
```

Compared with the full model 

# 7.5 Stepwise Regression

We assess bidirectional stepwise regression, giving the model access to all predictors.

```{r}
trControl <- trainControl(method = "none")

# Stepwise selection 
model_stepwise <- train(
  formula_full,
  data = train_data,
  trControl = trControl,
  method = "glmStepAIC",
  direction = "both",
  trace = FALSE
)
lm_stepwise <- model_stepwise$finalModel
```


```{r}

lm_stepwise_r2 <- 1 - (lm_stepwise$deviance / lm_stepwise$null.deviance)
n <- nrow(train_data)
p <- length(coef(lm_stepwise)) - 1
lm_stepwise_adj_r2 <- 1 - (1 - lm_stepwise_r2) * (n - 1) / (n - p - 1)
summary(lm_stepwise)
cat("Train Adj R^2:", lm_stepwise_adj_r2, "\n")
cat("Train RMSE:", accuracy(lm_stepwise$fitted.values, train_data$Ozone_ppm)[, "RMSE"], "\n")

pred <- predict(lm_stepwise, test_data)
actual <- test_data$Ozone_ppm
ss_res <- sum((actual - pred)^2)
ss_tot <- sum((actual - mean(actual))^2)
test_r2 <- 1 - ss_res / ss_tot
n_test <- nrow(test_data)
test_adj_r2 <- 1 - (1 - test_r2) * (n_test - 1) / (n_test - p - 1)
cat("Test Adj R^2: ", test_adj_r2, "\n")
cat("Test RMSE: ", accuracy(pred, actual)[, "RMSE"], "\n")
```

This model drops 6 of the 20 values in predictors. We see that it keeps in both Wind2m and Wind10m, whicih are heavily correlated, despite dropping similarly correlated predictors.

Forward and backward are not shown, but had similar, but inferior performance. 

The regression results indicate that meteorological conditions such as solar radiation, wind speed, and dew point have statistically significant relationships with ozone levels. Solar radiation and warmer conditions tend to increase ozone formation, while precipitation and wind may reduce ozone concentrations by dispersing pollutants.

# 8. Seasonality and Fourier Time_Series Regression

In this section we use techniques that are better suited and specifically designed for time-series data.

## 8.1 Autocorrelation (ACF)

We first start by creating an ACF plot.

```{r autocorrelation}

ozone_train_ts <- ts(train_data$Ozone_ppm, frequency = 365)
Acf(ozone_train_ts, main = "ACF of Daily Ozone (Train: 2019–2020)")

```

A slowly decaying ACF and repeating structure suggests strong seasonality. For ozone, this is expected due to sunlight-driven chemistry and summer weather patterns.


## 8.2 Select Fourier complexity (K)

We use AIC to select the number of Fourier harmonics (K). Larger K captures more complex seasonal shape but risks overfitting.

```{r fourier}

 k_values <- 1:10
 aic_values <- sapply(k_values, function(k) {
   formula <- as.formula(paste("ozone_train_ts ~ trend + fourier(ozone_train_ts, K=", k, ") +",
                                paste(significant_vars, collapse = " + ")))
   fit <- tslm(formula, data = train_data)
   return(AIC(fit))
 })

  ggplot(data.frame(K = k_values, AIC = aic_values), aes(x = K, y = AIC)) +
   geom_line() +
   geom_point() +
   scale_x_continuous(breaks = k_values) +
   labs(title = "AIC vs Number of Fourier Terms (K)", x = "K", y = "AIC") +
   theme_minimal()
```
Based on the plot K=3 or 5 is ideal.

## 8.3 Fit Fourier + predictors 

This is the corrected out-of-sample evaluation. We:

- compute Fourier terms on the training series

- project Fourier terms forward into the test horizon using fourierf()

- forecast with forecast(..., xreg=...)

```{r}
K <- 3
h <- nrow(test_data)

# time series used to fit
ozone_train_ts <- ts(train_data$Ozone_ppm, frequency = 365)

to_matrix <- function(df) df %>% select(all_of(predictors)) %>% mutate(across(everything(), as.numeric)) %>% as.matrix()

train_df <- cbind(
  ozone = as.numeric(ozone_train_ts),
  fourier(ozone_train_ts, K = K),
  to_matrix(train_data)
) %>% as.data.frame()

colnames(train_df) <- gsub("-", "_", colnames(train_df))

test_df <- as.data.frame(cbind(
  fourierf(ozone_train_ts, K = K, h = h),
  to_matrix(test_data)
))
colnames(test_df) <- gsub("-", "_", colnames(test_df))
fourier_features <- c(
  paste0(rep(c("S", "C"), K), rep(1:K, each = 2), "_365"), # fourier terms
  "NO2_ppb_or_ppm", "SO2_ppb_or_ppm", "Wind2m_ms", "DewPoint_C",
  "Precip_mm", "DayOfYear"
)


formula_fourier <- as.formula(paste("ozone_train_ts ~ trend +", paste(fourier_features, collapse = " + ")))

fit_fourier <- tslm(formula_fourier, data = train_df)

fc <- forecast(fit_fourier, newdata = test_df, h = h)

# Results
print("Train:")
cat("Adj R²:", summary(fit_fourier)$adj.r.squared, "\n")

cat("RMSE:",accuracy(fit_fourier)[, "RMSE"], "\n")



print("Test:")

n_test <- length(test_data$Ozone_ppm)
p <- length(coef(fit_fourier)) - 1
ss_res <- sum((test_data$Ozone_ppm - fc$mean)^2)
ss_tot <- sum((test_data$Ozone_ppm - mean(test_data$Ozone_ppm))^2)
r2     <- 1 - ss_res/ss_tot
adj_r2 <- 1 - (1 - r2) * (n_test - 1) / (n_test - p - 1)
cat("Adj R²:", adj_r2, "\n")
cat("RMSE:",accuracy(fc$mean, test_data$Ozone_ppm)[, "RMSE"], "\n")

```

The Fourier model captured seasonal patterns in the and testing data, performing significantly better than the linear models.

## 8.4 ARIMA

We assess the feasibility of augmenting the Fourier model with ARIMA.

```{r}
# auto.arima on residuals 
res.arima <- auto.arima(fit_fourier$residuals)
summary(res.arima)

# Combined fitted values and forecasts
combined_train <- fit_fourier$fitted.values + res.arima$fitted
lm_pred_test   <- forecast(fit_fourier, newdata = test_df, h = h)$mean
res_pred_test  <- forecast(res.arima, h = h)$mean
final_pred     <- lm_pred_test + res_pred_test

# align train if auto.arima differenced (differencing loses first d observations)
d <- res.arima$arma[6]  # number of differences chosen
ozone_train_aligned <- tail(as.numeric(ozone_train_ts), length(combined_train))

n <- length(ozone_train_aligned)
p <- length(coef(fit_fourier)) + length(coef(res.arima)) - 1

ss_res_train <- sum((ozone_train_aligned - combined_train)^2)
ss_tot_train <- sum((ozone_train_aligned - mean(ozone_train_aligned))^2)
r2_train     <- 1 - ss_res_train/ss_tot_train
adj_r2_train <- 1 - (1 - r2_train) * (n - 1) / (n - p - 1)

n_test <- length(test_data$Ozone_ppm)
ss_res_test <- sum((test_data$Ozone_ppm - final_pred)^2)
ss_tot_test <- sum((test_data$Ozone_ppm - mean(test_data$Ozone_ppm))^2)
r2_test     <- 1 - ss_res_test/ss_tot_test
adj_r2_test <- 1 - (1 - r2_test) * (n_test - 1) / (n_test - p - 1)

cat("ARIMA order chosen:", arimaorder(res.arima), "\n")

cat("\nTrain:\n")
cat("Adj R²:", adj_r2_train, "\n")
cat("RMSE:  ", accuracy(combined_train, ozone_train_aligned)[, "RMSE"], "\n")

cat("\nTest:\n")
cat("Adj R²:", adj_r2_test, "\n")
cat("RMSE:  ", accuracy(final_pred, test_data$Ozone_ppm)[, "RMSE"], "\n")
```
From the AR(2) coefficient of 0.49, we have moderate positive autocorrelation in the Fourier residuals. This confirms that residuals weren't white noise.
For training, using AR(1) has a significant improvement on Adj R^2 and RMSE.

However the effect it has is extremely small on the test set. This is because our forecast window is 365 days, and its effects become almost insignificant within a few weeks, without residuals to guide it. While ARIMA would be  effective on a model that has a shorter forecast window, it is unsuited to this large of a forecast window.

# 9. classification Modeling (Logistic Regression)

## 9.1 Define classification target

We classify days as “Orange or worse” using the AQI indicator already in the dataset:

- AQI_AtLeast_Orange = 1 if AQI ≥ 50

While regression predicts ozone levels directly, classification models are useful for public health decision making because they identify days likely to exceed air quality thresholds.

```{r logit}
train_cls <- train_data %>%
  mutate(AQI_AtLeast_Orange = factor(AQI_AtLeast_Orange, levels = c(0, 1)))

test_cls <- test_data %>%
  mutate(AQI_AtLeast_Orange = factor(AQI_AtLeast_Orange, levels = c(0, 1)))

```

## 9.2 Logistic regression (interpretable classifier)
We use a small, interpretable set of predictors (similar to regression drivers).

```{r}

cls_vars <- c("NO2_ppb_or_ppm", "SO2_ppb_or_ppm", "Wind2m_ms",
              "DewPoint_C", "Precip_mm", "Solar_Wm2", "DayOfYear")

cls_formula <- as.formula(paste("AQI_AtLeast_Orange ~", paste(cls_vars, collapse = " + ")))
logit_model <- glm(cls_formula, data = train_cls, family = binomial())
summary(logit_model)

```

# 10. PCA (Dimension Reduction)

PCA is used here to reduce redundancy among correlated weather variables. We:

- standardize predictors

- compute PCs

- examine variance explained and loadings

- optionally refit regression using PCs

Many meteorological variables are highly correlated (for example temperature and solar radiation), which can destabilize regression coefficients. PCA allows us to replace correlated variables with orthogonal components.

## 10.1 Prepare PCA predictors

```{r}

#pca_vars <- c("Temp_C","TempMax_C","TempMin_C","Wind2m_ms","Wind10m_ms",
#              "RH_percent","DewPoint_C","Precip_mm","Pressure_kPa","Solar_Wm2")

pca_vars<-predictors[1:length(predictors)-1]

pca_train <- train_data %>%
  select(all_of(pca_vars)) %>%
  mutate(across(everything(), as.numeric))

pca_fit <- prcomp(pca_train, center = TRUE, scale. = TRUE)

# Scree plot
plot(pca_fit, type = "l", main = "Scree Plot (PCA)")
summary(pca_fit)

```

## 10.2 Interpret component loadings

```{r}

loadings <- pca_fit$rotation
head(loadings)

```
From the loadings table, we see that the first PC is primarily is concerned with Temperature and Relative Humidity The second PC is primarily influenced by wind. The third is influenced by a combination of Wind and Relative Humidity.

## 10.3 Regression using PCs
We use enough PCs to explain ~80-90% variance. This come out to 4 PCs.

```{r}

var_explained <- cumsum(pca_fit$sdev^2) / sum(pca_fit$sdev^2)
n_pc <- which(var_explained >= 0.85)[1]
n_pc

```

```{r}

train_pcs <- as.data.frame(pca_fit$x[, 1:n_pc, drop = FALSE])
train_pcs$Ozone_ppm <- train_data$Ozone_ppm

pca_model <- lm(Ozone_ppm ~ ., data = train_pcs)
summary(pca_model)

```

Evaluate on 2021 data:

```{r}

pca_test <- test_data %>%
  select(all_of(pca_vars)) %>%
  mutate(across(everything(), as.numeric))

test_scores <- predict(pca_fit, newdata = pca_test)[, 1:n_pc, drop = FALSE]
pred_pca <- predict(pca_model, newdata = as.data.frame(test_scores))

cat("PCA Regression (Test 2021): RMSE =", RMSE(pred_pca, test_data$Ozone_ppm),
    " MAE =", MAE(pred_pca, test_data$Ozone_ppm), "\n")

```



# 11. Model Perfomance Comparison and Interpretation

```{r model comparison}

model_results <- data.frame(
  Model = c("Linear Regression",
            "Reduced Regression",
            "Stepwise Regression",
            "Fourier Regression",
            "PCA Regression"),
  
  RMSE = c(
    rmse_lm,
    RMSE(pred_red, obs_red),
    RMSE(pred, actual),
    accuracy(fc$mean, test_data$Ozone_ppm)[2],
    RMSE(pred_pca, test_data$Ozone_ppm)
  ),
  
  MAE = c(
    mae_lm,
    MAE(pred_red, obs_red),
    MAE(pred, actual),
    accuracy(fc$mean, test_data$Ozone_ppm)[3],
    MAE(pred_pca, test_data$Ozone_ppm)
  )
)

model_results

ggplot(model_results, aes(x = Model, y = RMSE)) +
  geom_bar(stat="identity", fill="steelblue") +
  theme_minimal() +
  labs(title="Model Comparison (Test RMSE)",
       y="RMSE",
       x="")

```

Based on RMSE and MAE, the Fourier regression performed best on the 2021 test set with the lowest RMSE of 0.00601, followed by stepwise regression. The standard linear and reduced regressions performed similarly in the middle, while PCA regression had the highest error, suggesting that dimensionality reduction through principal components loses predictive information present in the original variables.



